Patterns of floral and chromosome evolution in Onagraceae: a fossil-calibrated supermatrix analysis

### 2014-05-07 manuscript.tex
minor copy edits

\begin{abstract}
Onagraceae is a cosmopolitan plant family in the order Myrtales comprised of 22 genera and 657 species.
I used a GenBank data-mining approach to construct an 11 gene supermatrix of 521 taxa in Onagraceae and Lythraceae.
Maximum likelihood and Bayesian inference phylogenetic analyses were performed, 
and divergence time estimates were calibrated using 5 fossils.
Chromosome number evolution was estimated using probabilistic models,
and ancestral states of petal color and petal number were reconstructed using maximum likelihood.
Correlated evolution between petal color and petal number was tested using
Bayesian stochastic character mapping.
The monophyly of all major clades in Onagraceae was supported,
and Onagraceae was estimated to have diverged from Lythraceae 109 Mya. 
The base chromosome number of Onagraceae was inferred to be $x=5$.
Petal color and petal number were found
to follow a pattern of correlated evolution ($p=0.00$), 
suggesting concerted shifts in floral traits may play an important role shaping diversification of Onagraceae. 
\end{abstract}

@@

Lythraceae was selected to use as an outgroup since previous molecular phylogenetic analyses
place it sister to Onagraceae \citep{conti1997, Sytsma2004}.
I downloaded all DNA sequences from GenBank release 200 PLN division and
performed an exhaustive all-by-all BLASTn \citep{blast} comparison of sequences in Onagraceae
and Lythraceae.
Using a BLASTn e-value of $1.0 \times 10^{-10}$ threshold and a sequence length
percent similarity cutoff of 0.5,
I constructed clusters of putative homologs using a single-linkage hierarchical clustering algorithm.
Subspecies names were removed from all sequences, and all but one sequence of each species was pruned from each cluster.
Clusters that were not phylogenetically informative ($< 4$ taxa) were discarded,
and each cluster was aligned using MUSCLE \citep{edgar2004muscle}. 
The alignments were concatenated by species, and any species that was not present in at least
two clusters was removed from the supermatrix.
The program written to data-mine GenBank and assemble the supermatrix 
is available as the open source Python module 
SUMAC \citep{sumac}.
SUMAC can be used to assemble
supermatrices for any taxonomic group recognized in GenBank, 
and is optimized to run on multicore processors and clusters by utilizing multiple parallel processes.

@@

\paragraph{Phylogeny and divergence time estimates}
The topologies of the ML and Bayesian phylogenies were identical for all major clades within Onagraceae, 
so only the Bayesian MCC tree (Figures \ref{posteriors} and \ref{genera}) is shown here.
All Onagraceae genera described in \citet{wagner2007revised} were recovered as monophyletic clades with posterior probabilities of $> 0.95$
except for sister genera \textit{Neoholmgrenia} and \textit{Camissoniopsis} (posterior = 0.31) (Figure \ref{posteriors}).
Onagraceae was found to diverge from Lythraceae at 109 Mya (Figure \ref{genera}). 
Divergence time estimates of other major clades and 95\% highest posterior density (HPD) intervals can be seen in Table \ref{times}.

@@

\begin{table*}
   \center
   \begin{adjustbox}{max width=\textwidth}
      \setlength{\tabcolsep}{20pt}
      \begin{tabular}{lllllll}
         \toprule
	 & \multicolumn{4}{c}{Number of Petals} \\
         & 2 & 4 & 5 & 6 \\ 
	 \cmidrule{2-5}
	 Petal Color \\
         \hspace{3 mm} Pink & -0.005 & .011 & $ns$ & $ns$ \\
	 \hspace{3 mm} Yellow & -0.008 & .021 & $ns$ & $ns$ \\
	 \hspace{3 mm} White & 0.008 & 0.013 & -0.006 & -0.006 \\
	 \hspace{3 mm} Green & $ns$ & -0.011 & $ns$ & $ns$ \\
	 \hspace{3 mm} Red & $ns$ & $ns$ & $ns$ & $ns$ \\
         \bottomrule
      \end{tabular}
   \end{adjustbox}
   \caption{Test statistics for the correlation between petal color evolution and petal number evolution. 
	    $d_{ij}$ values are shown
            for the pairwise comparison of states with $p< 0.01$. $ns$ indicates no significant
	    association. The overall $D$ value was 0.263 ($p=0.00$).
	    }
   \label{correlations}
\end{table*}

>>>>

\begin{abstract}
Onagraceae is a cosmopolitan plant family in the order Myrtales comprised of 22 genera and 657 species.
I used a GenBank data-mining approach to construct an 11 gene supermatrix of 521 taxa in Onagraceae and Lythraceae.
Maximum likelihood and Bayesian inference phylogenetic analyses were performed, 
and divergence time estimates were calibrated using 5 fossils.
Chromosome number evolution was inferred using probabilistic models,
and ancestral states of petal color and petal number were reconstructed using maximum likelihood.
Correlated evolution between petal color and petal number was tested using
Bayesian stochastic character mapping.
The monophyly of all major clades in Onagraceae was supported,
and Onagraceae was estimated to have diverged from Lythraceae 109 Mya. 
The base chromosome number of Onagraceae was inferred to be $x=5$.
Petal color and petal number were found
to follow a pattern of correlated evolution ($p=0.00$), 
suggesting concerted shifts in floral traits may play an important role in shaping diversification of Onagraceae. 
\end{abstract}

@@

Lythraceae was selected as an outgroup since previous molecular phylogenetic analyses
place it sister to Onagraceae \citep{conti1997, Sytsma2004}.
I downloaded all DNA sequences from GenBank release 200 PLN division and
performed an exhaustive all-by-all BLASTn \citep{blast} comparison of sequences in Onagraceae
and Lythraceae.
Using a BLASTn e-value of $1.0 \times 10^{-10}$ threshold and a sequence length
percent similarity cutoff of 0.5,
I constructed clusters of putative homologs using a single-linkage hierarchical clustering algorithm.
Subspecies names were removed from all sequences, and all but one sequence of each species was pruned from each cluster.
Clusters that were not phylogenetically informative ($< 4$ taxa) were discarded,
and each cluster was aligned using MUSCLE \citep{edgar2004muscle}. 
The alignments were concatenated by species, and any species that was not present in at least
two clusters was removed from the supermatrix.
The program written to data-mine GenBank and assemble the supermatrix 
is available as the open source Python module 
SUMAC \citep{sumac}.
SUMAC will assemble
supermatrices for any taxonomic group recognized in GenBank, 
and is optimized to run on multicore processors and clusters by utilizing multiple parallel processes.

@@

\paragraph{Phylogeny and divergence time estimates}
The topologies of the ML and Bayesian phylogenies were identical for all major clades within Onagraceae, 
so only the Bayesian MCC tree (Figures \ref{posteriors} and \ref{genera}) is shown here.
All Onagraceae genera described in \citet{wagner2007revised} were recovered as monophyletic clades with posterior probabilities $> 0.95$
except for sister genera \textit{Neoholmgrenia} and \textit{Camissoniopsis} (posterior = 0.31) (Figure \ref{posteriors}).
Onagraceae was found to diverge from Lythraceae at 109 Mya (Figure \ref{genera}). 
Divergence time estimates of other major clades and 95\% highest posterior density (HPD) intervals can be seen in Table \ref{times}.

@@

\begin{table*}
   \center
   \begin{adjustbox}{max width=\textwidth}
      \setlength{\tabcolsep}{20pt}
      \begin{tabular}{lllllll}
         \toprule
	 & \multicolumn{4}{c}{Number of Petals} \\
         & 2 & 4 & 5 & 6 \\ 
	 \cmidrule{2-5}
	 Petal Color \\
         \hspace{3 mm} Pink & -0.005 & .011 & $ns$ & $ns$ \\
	 \hspace{3 mm} Yellow & -0.008 & .021 & $ns$ & $ns$ \\
	 \hspace{3 mm} White & 0.008 & 0.013 & -0.006 & -0.006 \\
	 \hspace{3 mm} Green & $ns$ & -0.011 & $ns$ & $ns$ \\
	 \hspace{3 mm} Red & $ns$ & $ns$ & $ns$ & $ns$ \\
         \bottomrule
      \end{tabular}
   \end{adjustbox}
   \caption{Test statistics for the correlation between petal color evolution and petal number evolution. 
	    $d_{ij}$ values are shown
            for the pairwise comparison of states with $p< 0.01$. $ns$ indicates no significant
	    association. Negative $d_{ij}$ values express negative correlation.
	    The overall $D$ value was 0.263 ($p=0.00$).
	    }
   \label{correlations}
\end{table*}
